Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-14 paper/paper.tex
last tweaks

% This project is part of the stellarTwins project.
% Copyright 2017 the authors.

@@

Projects like \project{The Cannon} (\citealt{ness15}; \citealt{casey17}; \citealt{ho17})
and \project{Avast} (Bedell et al.\ in preparation),
explore the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being asked.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable;
the model presented here has these same properties.
The \gaia\ data set is ideal for thinking about these kinds of purely statistical
models of stars.
The different stars observed by \gaia\ are measured at very different
signal-to-noise ratios,
and a data-driven model of the stellar color--magnitude distribution can capitalize
on the precise information coming from the high signal-to-noise stars
and deliver valuable information about the low signal-to-noise stars.
This information can then deliver precise parallax (or distance or absolute magnitude) inferences for all of the stars.

@@

\begin{eqnarray}
p(\varpi_{\true} \given \data, \hat{\alpha}) \, &\propto& \, \left|\frac{\dd\truth}{\dd\varpi}\right| \, p(\truth \given \data, \hat{\alpha}) \, f(\varpi_{\true}) \\
%p(\truth_0 \given \data, \hat{\alpha}) &\rightarrow& p(\varpi_{\true} \given y, \hat{\alpha}) \, f(\varpi_{\true}) \\
\mathrm{where} \; f(\varpi_{\true}) &=& \begin{cases}
              1 \quad \mathrm{if} \quad -2 < \mathrm{log}_{10} \varpi_{\true} < 2\\
              0 \quad \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}
with $f(\varpi_{\true})$ being a window function to insure $\varpi_{\true}$ is positive and to put $\varpi_{\true}$ on a similar grid for all the posteriors.

@@

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \tgas\ catalog (the y-axis of Figure~\ref{fig:delta}). This more directly shows what was alluded to in Figure~\ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \tgas\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
The precision of our posterior PDFs is greater than or equal to the precision of the \tgas\ catalogue for $87\%$ of the stars, shown as the dotted line at $x=0$,
and $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, which corresponds to a fraction decrease in variance by a factor of $>4$, shown as the dotted line at $y=0.14$.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{deltaCDF.pdf}
\caption{Fractional Change in Variance: The cumulative distribution function of the natural log of the fractional variance change when comparing our posterior to the \tgas\ catalogue. Negative numbers represent stars which have a posterior that is narrower than the \tgas\ catalogue, which is true for a large fraction of the stars. We've drawn three dashed lines. One represents the median star $(y=0.5)$ which has a fractional variation of $x=-0.3=ln(1/1.2^2)$, corresponding to an increase in precision by a factor of 1.2. Another represents the fraction of stars $(y=0.14)$ which have a fractional variance greater than $x=-1.4=ln(1/2^2)$, corresponding to an increase in precision by a factor of 2. The final line, $x=0$, represents the fraction of stars with an increased precision, $87\%$.}
\label{fig:deltaCDF}
\end{figure}

>>>>

% This project is part of the photoParallax project.
% Copyright 2017 the authors.

@@

Projects like \project{The Cannon} (\citealt{ness15}; \citealt{casey17}; \citealt{ho17})
and \project{Avast} (Bedell et al.\ in preparation),
explore the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being asked.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable;
the model presented here has these same properties.
Yet, the \gaia\ data set is ideal for thinking about these kinds of purely statistical
models of stars.
The different stars observed by \gaia\ are measured at very different
signal-to-noise ratios,
and a data-driven model of the stellar color--magnitude distribution can capitalize
on the precise information coming from the high signal-to-noise stars
and deliver valuable information about the low signal-to-noise stars.
This information can then deliver precise parallax (or distance or absolute magnitude) inferences for all of the stars.

@@

\begin{eqnarray}
p(\varpi_{\true} \given \data, \hat{\alpha}) \, &\propto& \, \left|\frac{\dd\truth}{\dd\varpi}\right| \, p(\truth \given \data, \hat{\alpha}) \, f(\varpi_{\true}) \\
\mathrm{where} \; f(\varpi_{\true}) &=& \begin{cases}
              1 \quad \mathrm{if} \quad -2 < \mathrm{log}_{10} \varpi_{\true} < 2\\
              0 \quad \mathrm{otherwise}
              \end{cases}
\label{eq:parallaxPost}
\end{eqnarray}
with $f(\varpi_{\true})$ being a window function to insure $\varpi_{\true}$ is positive and to put $\varpi_{\true}$ on a similar grid for all the posteriors.

@@

Figure~\ref{fig:deltaCDF} shows the cumulative distribution function of the natural log of the fractional change in variance of the parallax posterior pdf relative to the \tgas\ catalog (the y-axis of Figure~\ref{fig:delta}). This more directly shows what was alluded to in Figure~\ref{fig:posterior}. For the majority of the \gaia\ stars, our prior slightly increases the precision of the posterior PDF compared with the \tgas\ variance. Our parallax posterior PDFs are more precise than the \tgas\ catalog entries by a median factor of $1.2$, shown as the dotted line at $y=0.5$.
The precision of our posterior PDFs is greater than or equal to the precision of the \tgas\ catalogue for $83\%$ of the stars, shown as the dotted line at $x=0$,
and $14\%$ of our parallax posterior PDFs are more precise than the \tgas\ catalog by a factor of $>2$, which corresponds to a fraction decrease in variance by a factor of $>4$, shown as the dotted line at $y=0.14$.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{deltaCDF.pdf}
\caption{Fractional Change in Variance: The cumulative distribution function of the natural log of the fractional variance change when comparing our posterior to the \tgas\ catalogue. Negative numbers represent stars which have a posterior that is narrower than the \tgas\ catalogue, which is true for a large fraction of the stars. We've drawn three dashed lines. One represents the median star $(y=0.5)$ which has a fractional variation of $x=-0.3=ln(1/1.2^2)$, corresponding to an increase in precision by a factor of 1.2. Another represents the fraction of stars $(y=0.14)$ which have a fractional variance greater than $x=-1.4=ln(1/2^2)$, corresponding to an increase in precision by a factor of 2. The final line, $x=0$, represents the fraction of stars with an increased precision, $83\%$.}
\label{fig:deltaCDF}
\end{figure}
